\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-27 paper.qmd
Added Figure 7 to choose te number of bins

```{r}
#| label: hexbin-two-curvy
## Compute hexbin parameters
num_bins_x_two_curvy <- 6

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_two_curvy_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5, size = 0.5, color = "#000000") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)
```

@@

@@

@@

@@

>>>>

```{r}
#| label: hexbin-two-curvy
## Compute hexbin parameters
num_bins_x_two_curvy <- 7

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_two_curvy_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5, size = 0.5, color = "#000000") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
             aes(x = UMAP1, y = UMAP2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("a", position = c(0.92, 0.93)) +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.35, 2))

@@

```{r}
#| label: hexbin-regular-two-curvy2
## hexagon binning to have regular hexagons
hb_obj_two_curvy2 <- hex_binning(
  data = umap_two_curvy_scaled, 
  bin1 = 10, 
  r2 = r2)

a1_2 <- calc_bins_y(
  bin1 = 10, 
  r2 = r2
  )$a1

## Data set with all centroids
all_centroids_df2 <- hb_obj_two_curvy2$centroids

## Generate all coordinates of hexagons
hex_grid2 <- hb_obj_two_curvy2$hex_poly

## To obtain the standardise counts within hexbins
counts_df2 <- hb_obj_two_curvy2$std_cts
df_bin_centroids2 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df2, 
  counts_df = counts_df2) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 11")

hex_grid_with_counts_s_curve2 <- full_join(
  hex_grid2, 
  df_bin_centroids2 |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.35, 2)) +
  interior_annotation("b")
```

```{r}
#| label: hexbin-regular-two-curvy3
## hexagon binning to have regular hexagons
hb_obj_two_curvy3 <- hex_binning(
  data = umap_two_curvy_scaled, 
  bin1 = 13, 
  r2 = r2)

a1_3 <- calc_bins_y(
  bin1 = 13, 
  r2 = r2
)$a1

## Data set with all centroids
all_centroids_df3 <- hb_obj_two_curvy3$centroids

## Generate all coordinates of hexagons
hex_grid3 <- hb_obj_two_curvy3$hex_poly

## To obtain the standardise counts within hexbins
counts_df3 <- hb_obj_two_curvy3$std_cts
df_bin_centroids3 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df3, 
  counts_df = counts_df3) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 14")

hex_grid_with_counts_s_curve3 <- full_join(hex_grid3, 
                                          df_bin_centroids3 |> select(hexID, std_counts), 
                                          by = c("hex_poly_id" = "hexID"))

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.35, 2)) +
  interior_annotation("c")
```

@@

```{r}
#| label: error-two-curvy

error_two_curvy_clust_umap <- read_rds("data/two_curvy_clust/error_two_curvy_clust_umap.rds")

## Find the minimum MSE when have duplicate a1
error_two_curvy_clust_umap <- error_two_curvy_clust_umap |>
  group_by(a1) |>
  filter(MSE == min(MSE)) |>
  ungroup() |>
  mutate(prop_dens = b_non_empty/side_length)

base_line_prop_dens <- error_two_curvy_clust_umap |>
  filter(a1 == min(a1)) |>
  pull(prop_dens)

error_two_curvy_clust_umap <- error_two_curvy_clust_umap |>
  mutate(prop_comp = prop_dens/base_line_prop_dens)

mse_two_curvy_b <- ggplot(error_two_curvy_clust_umap, 
                     aes(x = a1, 
                         y = MSE)) +
  geom_vline(xintercept = 0.10, linetype=2,
           color = "#54278f", linewidth=1) +
  geom_vline(xintercept = 0.12,
             linetype="solid",
             color = "#006d2c",
             linewidth=1) +
  geom_vline(xintercept = 0.18, linetype=2,
             color = "#08519c", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 3, 5, 6, 8, 10:12)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "MSE") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

prop_dens_a1 <- ggplot(error_two_curvy_clust_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.12, linetype="solid",
             color = "#006d2c", linewidth=1) +
  geom_vline(xintercept = 0.10, linetype=2,
             color = "#54278f", linewidth=1) +
  geom_vline(xintercept = 0.18, linetype=2,
             color = "#08519c", linewidth=1) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 3, 5, 6, 8, 10:12)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("proportion of densities")) +
  ggtitle("(d)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

error_two_curvy_clust_umap <- error_two_curvy_clust_umap |>
  mutate(prop_bins = b_non_empty/b)

a1_m_two_curvy <- ggplot(error_two_curvy_clust_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.12, linetype="solid",
             color = "#006d2c", linewidth=1) +
  geom_vline(xintercept = 0.10, linetype=2,
             color = "#54278f", linewidth=1) +
  geom_vline(xintercept = 0.18, linetype=2,
             color = "#08519c", linewidth=1) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 3, 5, 6, 8, 10:12)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("proportion of non-empty bins ", bgroup("(", frac(m, b), ")")))) +
  ggtitle("(b)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

@@

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the S-curve data, using three different bin inputs: (a) $b = 91 \\text{ } (7, \\text{ }13)$, (b) $b = 220 \\text{ } (11, \\text{ }20)$, and (c) $b = 312 \\text{ } (13, \\text{ }24)$. Color indicates standardized counts, dark indicating high count and light indicates low count. At the smallest bin size the data segregates into two separate groups, suggesting this is too many bins. Using the MSE of the model fit in $p\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 6
#| fig-height: 4

# hex_grid_coloured_two_curvy1 + 
#   hex_grid_coloured_two_curvy2 + 
#   hex_grid_coloured_two_curvy3 +
#   plot_layout(guides='collect', ncol = 3,
#               heights = c(2,1)) &
#   theme(legend.position='none', plot.tag = element_text(size = 8))

hex_grid_coloured_two_curvy1 +
  hex_grid_coloured_two_curvy2 +
  hex_grid_coloured_two_curvy3 +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

```{r}
#| label: rm-lwd-bin-error

error_rm_two_curvy <- read_rds("data/two_curvy_clust/error_rm_lwd_diff_bin.rds") |>
  mutate(bin1  = as.factor(bin1))

mse_two_curvy_lwd <- ggplot(error_rm_two_curvy, 
                     aes(x = benchmark_rm_lwd, 
                         y = MSE,
                         color = bin1)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.3
    ) + 
  # geom_vline(xintercept = benchmark1, linetype="solid", 
  #            color = "#bdbdbd", linewidth=1, alpha = 0.5) +
  # scale_x_continuous("standardized bin count", 
  #        transform = "reverse") + 
  scale_x_continuous(breaks = unique(error_rm_two_curvy$benchmark_rm_lwd)) +
  scale_color_manual(values=c("#08519c", "#006d2c", "#54278f")) +
  xlab("threshold to remove low-density hexagons") +
  ylab("MSE") +
  ggtitle("(c)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none")
```

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best hexagon bin parameters, thresholds to remove low density bins and large edges. Both (b) and (c) show MSE, against number of bins along the x-axis and standardised count. A good benchmark value for these parameters is when the MSE drops and then flattens out. Plot (a) shows the distribution of stadardised counts of hexagons. Plot (d) shows the distribution of $2\\text{-}D$ Euclidean distances between bin centroids, with a good benchmark value for removing large edges would being the distance that shows the first large decrease."
#| label: fig-param-two-curvy
#| out-width: 100%
#| fig-width: 12
#| fig-height: 10
#| fig-pos: H

mse_two_curvy_b + a1_m_two_curvy +
  mse_two_curvy_lwd + prop_dens_a1 +
  plot_layout(ncol=2)
```

### 2024-09-27 paper.tex
Added Figure 7 to choose te number of bins

@@

>>>>

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-bins-two-curvy-1.pdf}

\caption{\label{fig-bins-two-curvy}Hexbin density plots of UMAP layout
of the S-curve data, using three different bin inputs: (a)
\(b = 91 \text{ } (7, \text{ }13)\), (b)
\(b = 220 \text{ } (11, \text{ }20)\), and (c)
\(b = 312 \text{ } (13, \text{ }24)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. At the
smallest bin size the data segregates into two separate groups,
suggesting this is too many bins. Using the MSE of the model fit in
\(p\text{-}D\) helps decide on a useful choice of number of bins.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-param-two-curvy-1.pdf}

\caption{\label{fig-param-two-curvy}Various plots to help assess best
hexagon bin parameters, thresholds to remove low density bins and large
edges. Both (b) and (c) show MSE, against number of bins along the
x-axis and standardised count. A good benchmark value for these
parameters is when the MSE drops and then flattens out. Plot (a) shows
the distribution of stadardised counts of hexagons. Plot (d) shows the
distribution of \(2\text{-}D\) Euclidean distances between bin
centroids, with a good benchmark value for removing large edges would
being the distance that shows the first large decrease.}
